### 2021-01-03 HW-AdvancedQFT2020/advqft1/advqft1.tex
Update advqft notes, add YM2 notes
... include more details about:
    - quantum charges
    - FP quantization

	Conserved charge $
		Q = \int\dd[2]{x} j^0(x)
	$, we have:
	\begin{equation}
		P_\mu = - \int\dd{x^1} \pd^0 X_\mu
		= \int\dd{x^1} \pd_0 X_\mu,\quad
		M_{\mu\nu}
		= \frac{1}{2}\,\int\dd{x^1}\pqty{
			X_\nu\,\pd_0 X_\mu
			- X_\mu\,\pd_0 X_\nu
		}
	\end{equation}
	They can be interpreted as spacetime momentum and spacetime angular momentum. 
	\qedfull
	\end{itemize}
\pagebreak[3]
\subsection{Real Scalar in $(3+1)$\,D}
	\vspace{-.8\baselineskip}
	\begin{equation}
		\mcal{L}
		= -\frac{1}{2}\,
			\pd^\mu\phi\,
			\pdd{\mu}\phi
		-\frac{1}{2}\,m^2\phi^2
	\end{equation}
	\begin{itemize}
	\item For $\phi$: scalar, under $
		x' = \lambda\circ x
	$, $\phi(x)\mapsto \phi'(x)$, while:
	\begin{equation}
		\phi'(x') = \phi(x)
		\ \Longrightarrow\ %
		\phi'(x) = \phi(\lambda^{-1}\circ x)
	\end{equation}
	For $\lambda\sim \lambda\id{^\mu_\nu}$: Lorentz transformation, $
		\eta_{\mu\nu}
			\lambda\id{^\mu_\rho}
			\lambda\id{^\nu_\sigma}
		= \eta_{\rho\sigma}
	$, or equivalently, $
		(\lambda^{-1})\id{^\mu_\nu}
		= \lambda\id{_\nu^\mu}
	$. Therefore,
	\begin{equation}
		\phi'(x^\mu) = \phi(\lambda^{-1}\circ x^\mu)
		= \phi(x^\nu\lambda\id{_\nu^\mu})
	\end{equation}
\pagebreak[3]

@@

	Back to our Lorentz transformation $
		\var{\phi}
		= -\omega\id{^\mu_\nu} x^\nu\pdd{\mu}\phi
	$, we have symmetry variation:
	\begin{equation}
		\var{\mcal{L}}
		= - \omega\id{^\mu_\nu} x^\nu
			\pdd{\mu} \mcal{L}
		= - \pdd{\mu} \pqty{
			\omega\id{^\mu_\nu} x^\nu
			\mcal{L}
		}
	\end{equation}
	This gives a boundary term $
		K^\mu = -\omega\id{^\mu_\nu} x^\nu \mcal{L}
	$, and the Noether current and its corresponding conserved charge can be calculated as follows:
	\begin{gather}
		j^\mu = -\omega\id{^\sigma_\nu} x^\nu
		\pqty{
			\pdv{\mcal{L}}{(\pdd{\mu}\phi)}
				\,\pdd{\sigma} \phi
			- \delta^\mu_\sigma \mcal{L}
		},\\[1ex]
		Q = \int \dd[3]{x} j^0
		= - \omega\id{^\sigma_\nu} \int \dd[3]{x}
		x^\nu \pqty{
			\pdd{0}\phi\,\pdd{\sigma}\phi
			- \delta^0_\sigma \mcal{L}
		},
	\end{gather}

	Note that $\omega\id{^\mu_\nu}$ is arbitrary, therefore $Q$ can be decomposed into independent charges:
	\begin{equation}
		Q = \omega_{\mu\nu} M^{\mu\nu},\quad
		M^{\mu\nu}
		= - \int \dd[3]{x}
		x^{[\mu} \pqty{
			\pdd{0}\phi\,\pd^{\nu]}\phi
			- \eta^{\nu] 0} \mcal{L}
		},
	\end{equation}
	The indices of $M^{\mu\nu}$ are anti-symmetrized to match the degrees of freedom in $\omega_{\mu\nu}$. Note that the $\mcal{L}$ term only appears when one of the indices is 0.

	Canonical quantization:
	\begin{gather}
		\dot{\phi} = \pdd{0}\phi,\quad
		\Pi = \pdv{\mcal{L}}{\dot{\phi}}
		= \dot{\phi},\quad
		[\phi(\vb{x}),\Pi(\vb{y})]
		= [\phi,\dot{\phi}](\vb{x})
		= i\delta(\vb{x} - \vb{y})
	\end{gather}
	Other equal-time commutators between $\phi,\Pi$ all just vanish. We have:
	\begin{equation}
		M^{0i} = -M^{i0}
		= -\frac{1}{2}\,
		\int \dd[3]{x} \pqty{
			\dot{\phi}\,\pqty{
				x^0 \pd^i
				- x^i \pd^0
			}\,\phi
			+ x^i \mcal{L}
		},\\
		M^{ij}
		= -\frac{1}{2}\,\int \dd[3]{x}
			\dot{\phi}\,\pqty{
				x^i \pd^j
				- x^j \pd^i
			}\,\phi
	\end{equation}

	Notice that $
		x^{[\mu} \pd^{\nu]}
		= \frac{1}{2}\,\pqty{
			x^\mu \pd^\nu - x^\nu \pd^\mu
		} = \frac{1}{2} D^{ij}
	$ is the Killing vector fields of $\mbb{R}^{3,1}$, hence they naturally follow the commutation relations of $\mfrak{so}(3,1)$ (up to a constant coefficient)\footnote{
		I would like to thank \textit{林般} for pointing this out. 
	}. We have:
	\begin{equation}
	\begin{aligned}
		[M^{ij}, M^{kl}]
		&= \frac{1}{4}\,\int \dd[3]{x}
			\int \dd[3]{y} \bqty{
				\dot{\phi} D^{ij}\phi(x),
				\dot{\phi} D^{kl}\phi(y)
			} \\
		&= \frac{1}{4}\,\int \dd[3]{x}
			\dot{\phi}\,\bqty{
				 D^{ij}, D^{kl}
			}\,\phi
	\end{aligned}
	\end{equation}
	Similar holds for $M^{i0}$. Therefore, $M^{\mu\nu}$'s indeed form the Lie algebra $\mfrak{so}(3,1)$. 
	\qedfull

>>>>

	Conserved charge $
		Q = \int\dd[2]{x} j^0(x)
	$, we have:
	\begin{equation}
		P_\mu = - \int\dd{x^1} \pd^0 X_\mu
		= \int\dd{x^1} \pd_0 X_\mu,\quad
		M_{\mu\nu}
		= \frac{1}{2}\,\int\dd{x^1}\pqty{
			X_\nu\,\pd_0 X_\mu
			- X_\mu\,\pd_0 X_\nu
		}
	\end{equation}
	They can be interpreted as spacetime momentum and spacetime angular momentum. 
	\qedfull
	\end{itemize}
%\pagebreak[3]
\subsection{Real Scalar in $(3+1)$\,D}
	\vspace{-.8\baselineskip}
	\begin{equation}
		\mcal{L}
		= -\frac{1}{2}\,
			\pd^\mu\phi\,
			\pdd{\mu}\phi
		-\frac{1}{2}\,m^2\phi^2
	\end{equation}
	\begin{itemize}
	\item For $\phi$: scalar, under $
		x' = \lambda\circ x
	$, $\phi(x)\mapsto \phi'(x)$, while:
	\begin{equation}
		\phi'(x') = \phi(x)
		\ \Longrightarrow\ %
		\phi'(x) = \phi(\lambda^{-1}\circ x)
	\end{equation}
	For $\lambda\sim \lambda\id{^\mu_\nu}$: Lorentz transformation, $
		\eta_{\mu\nu}
			\lambda\id{^\mu_\rho}
			\lambda\id{^\nu_\sigma}
		= \eta_{\rho\sigma}
	$, or equivalently, $
		(\lambda^{-1})\id{^\mu_\nu}
		= \lambda\id{_\nu^\mu}
	$. Therefore,
	\begin{equation}
		\phi'(x^\mu) = \phi(\lambda^{-1}\circ x^\mu)
		= \phi(x^\nu\lambda\id{_\nu^\mu})
	\end{equation}
%\pagebreak[3]

@@

	Back to our Lorentz transformation $
		\var{\phi}
		= -\omega\id{^\mu_\nu} x^\nu\pdd{\mu}\phi
	$, we have symmetry variation:
	\begin{equation}
		\var{\mcal{L}}
		= - \omega\id{^\mu_\nu} x^\nu
			\pdd{\mu} \mcal{L}
		= - \pdd{\mu} \pqty{
			\omega\id{^\mu_\nu} x^\nu
			\mcal{L}
		}
	\end{equation}
	We can write this down without explicit calculations, since we know $\mcal{L}$ itself is a Lorentz scalar, and that's how scalar transforms under Lorentz transformations. \\
	This gives a boundary term $
		K^\mu = -\omega\id{^\mu_\nu} x^\nu \mcal{L}
	$, and the Noether current and its corresponding conserved charge can be calculated as follows:
	\begin{gather}
		j^\mu = -\omega\id{^\sigma_\nu} x^\nu
		\pqty{
			\pdv{\mcal{L}}{(\pdd{\mu}\phi)}
				\,\pdd{\sigma} \phi
			- \delta^\mu_\sigma \mcal{L}
		},\\[1ex]
		Q = \int \dd[3]{x} j^0
		= - \omega\id{^\sigma_\nu} \int \dd[3]{x}
		x^\nu \pqty{
			\pdd{0}\phi\,\pdd{\sigma}\phi
			- \delta^0_\sigma \mcal{L}
		},
	\end{gather}

	Note that $\omega\id{^\mu_\nu}$ is arbitrary, therefore $Q$ can be decomposed into independent charges:
	\begin{equation}
		Q
		= \frac{1}{2}\,
			\omega_{\mu\nu} M^{\mu\nu},\quad
		M^{\mu\nu}
		= - \int \dd[3]{x}
			2x^{[[\mu} \pqty{
				\pdd{0}\phi\,\pd^{\nu]}\phi
				- \eta^{\nu] 0} \mcal{L}
			},
	\end{equation}
	The indices of $M^{\mu\nu}$ are anti-symmetrized to match the degrees of freedom in $\omega_{\mu\nu}$. Note that the $\mcal{L}$ term only appears when one of the indices is 0.

	Note that the canonical momentum:
	\begin{equation}
		\varpi
		= \pdv{\mcal{L}}{\dot{\phi}}
		= \dot{\phi}
		= \pd_0\phi
	\end{equation}
	It is thus natural to re-organize $M^{0i}$ in the following way:
	\begin{equation}
	\begin{aligned}
		M^{0i} = -M^{i0}
		&= - \int \dd[3]{x} \pqty{
			\varpi\,\pqty{
				x^0 \pd^i
				- x^i \pd^0
			}\,\phi
			- x^i \mcal{L}
		} \\
		&= - \int \dd[3]{x} \pqty{
			x^0 \varpi\,\pd^i \phi
			+ x^i \pqty\big{
				\varpi\dot{\phi} - \mcal{L}
			}
		} \\
		&= - \int \dd[3]{x} \pqty{
			x^0 \varpi\,\pd^i \phi
			+ x^i \mcal{H}
		} \\
	\end{aligned}
	\end{equation}

	We've obtained an interesting result: the expression for the boost generator $M^{i0}$ contains the Hamiltonian density $\mcal{H}$, weighted by the radial distance $x^i$. This is natural since a boost does indeed contains time evolution for states away from the origin. It's an important result utilized by the so-called \textit{Rindler decomposition}; in fact, $M^{i0}$ becomes the Hamiltonian for an accelerated observer in the Rindler patch\footnote{
		See the lecture notes of Tom Hartman: \http{hartmanhep.net/topics2015/gravity-lectures.pdf} or Daniel Harlow \arxiv{1409.1231}. 
	}.

	For $M^{ij}$, we have:
	\begin{equation}
		M^{ij}
		= - \int \dd[3]{x}
			\dot{\phi}\,\pqty{
				x^i \pd^j
				- x^j \pd^i
			}\,\phi
	\end{equation}
	This is interpreted as the angular momentum of the field $\phi$. Suppose $\phi$ is a wave packet localized around $\vb{x}$ with momentum $\approx \vb{p}$, then we have the classical angular momentum up to some factor:
	\begin{equation}
		M^{ij}
		\sim \pqty{
				x^i p^j
				- x^j p^i
			} \int \dd[3]{x}
				E \phi^2,
	\quad
		E = \sqrt{\vb{p}^2 + m^2}
	\end{equation}

	The $
		\int \dd[3]{x} E \phi^2
	$ factor in the above expression is an $\order{1}$ normalization constant for a particle-like wave packet; to see this, note that $\phi\in\mbb{R}$ has a phase factor $
		\phi
		\sim a\,e^{+ik\cdot x} + a^\dagger e^{-ik\cdot x}
		\sim \cos\,(k\cdot x)
	$,
	\begin{gather}
	\small
		E = \int \dd[3]{x} \mcal{H}
		= \int \dd[3]{x} \pqty{
				\frac{1}{2} \dot{\phi}^2
				+ \frac{1}{2} (\grad{\phi})^2
				+ \frac{1}{2} m^2 \phi^2
				+ \cdots
			}
		\sim \pqty{E^2 + \vb{p^2} + m^2 + \cdots}
			\int \dd[3]{x} \frac{1}{2}\,\phi^2,
	\end{gather}
	\vspace{-.7\baselineskip}
	\begin{equation}
		E\int \dd[3]{x} \phi^2 \sim 1,
	\end{equation}
	Indeed, we have: $
		M^{ij}
		\sim \pqty{
				x^i p^j
				- x^j p^i
			}
	$.

	Canonical quantization:
	\begin{gather}
		[\phi(\vb{x}),\varpi(\vb{y})]
		= [\phi(\vb{x}),\dot{\phi}(\vb{y})]
		= i\delta(\vb{x} - \vb{y})
	\end{gather}
	Other equal-time commutators between $\phi,\varpi$ all just vanish. Operator products are then regularized by normal ordering: $M\mapsto \normorder{M}$, which can be explicitly implemented by normal ordering of the oscilator modes:
	\begin{equation}
		\phi(x)
		= \int \frac{\dd[3]{k}}{(2\pi)^3}
			\frac{1}{\sqrt{2E_k}}
			\pqty{
				a_{\vb{k}} e^{ik\cdot x}
				+ a^\dagger_{\vb{k}'} e^{-ik\cdot x}
			},
	\quad
		[a_{\vb{k}}, a^\dagger_{\vb{k}'}]
		= (2\pi)^3 \delta(\vb{k} - \vb{k}')
	\end{equation}

	For example, the first term in $M^{0i} = -M^{i0}$ can be expanded as:
	\begin{equation}
	\begin{aligned}
		- x^0 \int \dd[3]{x}
			\varpi\,\pd^i \phi
		= x^0 \int \frac{\dd[3]{k}}{(2\pi)^3}\,
				k^i a^\dagger_k a_k
		= x^0 P^i
	\end{aligned}
	\end{equation}
	Here $P^\mu$ is the momentum operator on the Hilbert space, promoted from the classical $-i\pd^\mu$. $M^{0i}$ is thus further reduced to:
	\begin{equation}
		M^{0i} = -M^{i0}
		= x^0 P^i - \int \dd[3]{x}
			x^i \mcal{H}
	\end{equation}

	We note that this result is almost the classical $
		x^0 P^i - x^i P^0
	$, but here $x^i P^0$ is replaced with the integral over energy density $\mcal{H}$. The result can be nicely re-written with the stress tensor $T^{\mu\nu}$; just run the Noether's procedure with $\var{x^\mu} = \epsilon^\mu$, then we shall obtain:
	\begin{equation}
	\begin{gathered}
		j'^\mu
		= \epsilon_\nu T^{\mu\nu},
	\quad
		T^{\mu\nu}
		= \pd^\mu\phi\,\pd^\nu\phi
			+ \eta^{\mu\nu} \mcal{L},
	\\
		Q' = \epsilon_\mu P^\mu,
	\quad
		P^\mu
		= \int \dd[3]{x} T^{\mu 0}
		= \int \dd[3]{x} \pqty{
			\pd^0\phi\,\pd^\mu\phi
			+ \eta^{0\mu} \mcal{L}
		}
	\\
		M^{\mu\nu}
		= \int \dd[3]{x}
			2x^{[\mu} T^{\nu]0}
	\end{gathered}
	\end{equation}
	The quantization $M\mapsto \normorder{M}$ is thus reduced to the quantization of $T_{0\nu}$.

	\sidenote{TODO: Detailed analysis! HINT: Ward identity!}

	Notice that $
		x^{[\mu} \pd^{\nu]}
		= \frac{1}{2}\,\pqty{
			x^\mu \pd^\nu - x^\nu \pd^\mu
		} = \frac{1}{2} D^{\mu\nu}
	$ is the Killing vector fields of $\mbb{R}^{3,1}$, hence they naturally follow the commutation relations of $\mfrak{so}(3,1)$ (up to a constant coefficient, or an isomorphism)\footnote{
		I would like to thank \textit{林般} for pointing this out. 
	}. We have:
	\begin{equation}
	\begin{aligned}
		[M^{\mu\nu}, M^{\rho\sigma}]
		&= \int \dd[3]{x}
			\int \dd[3]{y} \bqty{
				\dot{\phi} D^{\mu\nu}\phi(x),
				\dot{\phi} D^{\rho\sigma}\phi(y)
			} \\
		&= \int \dd[3]{x}
			\dot{\phi}\,\bqty{
				 D^{\mu\nu}, D^{\rho\sigma}
			}\,\phi
	\end{aligned}
	\end{equation}
	Similar holds for $M^{i0}$. Therefore, $M^{\mu\nu}$'s indeed form the Lie algebra $\mfrak{so}(3,1)$. 
	\qedfull
